% AdaptEvolve-MAS — AI Safety Paper
% Target: SafeAI @ AAAI 2026 (4-6 pages) / AIES 2026
% Compile: pdflatex main.tex && bibtex main && pdflatex main.tex && pdflatex main.tex

\documentclass[letterpaper]{article}
\usepackage{aaai26}         % AAAI 2026 style
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{xcolor}
\usepackage{multirow}

\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}

\title{AdaptEvolve-MAS: Online Objective Correction in\\
       Meta-Evolutionary Multi-Agent Code Optimization}

\author{
  % Author names hidden for blind review
  % Pranamya Kulkarni, [Team Members] \\
  % GESCOEMSR, Nashik \\
  Anonymous Authors \\
  Submitted to SafeAI @ AAAI 2026
}

\begin{document}

\maketitle

% ─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
Closed-loop LLM-driven code optimization systems are vulnerable to three
failure modes: \emph{objective lock-in} (static fitness functions can be
gamed), \emph{open-loop search} (no mechanism to detect proxy divergence
from the true goal), and \emph{opaque adaptation} (the system modifies its
behaviour without an interpretable audit trail).
We present \textbf{AdaptEvolve-MAS}, a multi-agent meta-evolutionary framework
that addresses all three via four verifiable safety properties:
\textbf{Online Objective Correction} (OOC),
\textbf{Bounded Autonomy} (BA),
\textbf{Interpretable Decision Trails} (IDT), and
\textbf{Corrigible Restructuring} (CR).
The central mechanism is a \emph{Mechanic} agent that inspects
cross-cycle performance history and dynamically re-weights evaluation
criteria within a single inference run---without fine-tuning, without
external human feedback, and without halting the optimization loop.
We formalize OOC as an inference-time safety property, introduce a
\emph{Safety Scenario Suite} of four adversarial test cases, and report
ablation results on three benchmark coding tasks.
OOC improves holistic convergence scores over a no-meta-learning baseline
across all tasks while maintaining BA and IDT across all runs.
\end{abstract}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}

The recent wave of LLM-based evolutionary code optimization
systems~\cite{funSearch2023,openEvolve2024} demonstrates that language
models can act as mutation and evaluation operators within a search loop.
Yet these systems share a structural vulnerability: the fitness function
is defined once and remains \emph{fixed} throughout an optimization run.
This creates three compounding failure modes.

\textbf{(1) Objective lock-in.} A fixed proxy metric
(e.g., execution time) can be satisfied by degenerate solutions---a
program that immediately returns without computing anything scores
perfectly on speed but zero on correctness.  This is a form of
\emph{specification gaming}~\cite{krakovna2020}.

\textbf{(2) Open-loop search.} Without a mechanism that observes
\emph{whether the proxy is tracking the true goal}, the system has no
way to detect and correct divergence mid-run.

\textbf{(3) Opaque adaptation.} Even systems that adapt their search
strategy rarely expose why---the rationale for each change is not
recorded, making post-hoc auditing impossible.

We address all three with \textbf{AdaptEvolve-MAS}, a four-agent LangGraph
pipeline in which a specialized \emph{Mechanic} agent performs
\emph{Online Objective Correction}: it reads the cross-cycle performance
log, identifies proxy-goal misalignment, and proposes reweighted
evaluation criteria for the next cycle---all within a single forward
pass, with no gradient updates and no human labels required.

\noindent\textbf{Contributions:}
\begin{enumerate}
  \item Formalization of OOC as an inference-time AI safety property.
  \item The AdaptEvolve-MAS Safety Property Set: BA, IDT, OOC, CR.
  \item A Safety Scenario Suite (SS-1--SS-4) that adversarially probes each property.
  \item Ablation evidence that OOC improves convergence on three benchmark tasks.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Background and Related Work}

\paragraph{Specification gaming and reward hacking.}
\citet{krakovna2020} catalogue 60+ documented cases of agents satisfying
the letter but not the spirit of their objective.
\citet{amodei2016} identify reward misspecification as a core alignment
problem.  Our OOC mechanism directly targets this failure mode.

\paragraph{Constitutional AI and RLHF.}
\citet{bai2022} and \citet{ouyang2022} modify objectives via supervised
fine-tuning or preference learning \emph{before} or \emph{between}
deployments.  AdaptEvolve-MAS differs: OOC operates \emph{during} a single
inference run, requiring neither a labeled dataset nor parameter updates.

\paragraph{Corrigibility.}
\citet{soares2015} define corrigibility as the disposition to defer to
human oversight.  AdaptEvolve-MAS implements \emph{active} corrigibility:
the Mechanic \emph{self-initiates} objective corrections rather than
waiting for external intervention.

\paragraph{Evolutionary code optimization.}
FunSearch~\cite{funSearch2023} and OpenEvolve~\cite{openEvolve2024} use
LLMs as mutation operators within a fixed fitness landscape.  Neither
provides a second-order mechanism to update what the fitness function
measures mid-run.

\paragraph{Multi-agent frameworks.}
AutoGen~\cite{wu2023} and MetaGPT~\cite{hong2023} coordinate multiple
LLM agents but do not include an agent whose role is to restructure
what the other agents optimize \emph{for}.

% ─────────────────────────────────────────────────────────────────────────────
\section{The AdaptEvolve-MAS Framework}

\subsection{System Architecture}

Let the system state at cycle $c$ be
$\mathcal{S}_c = (\text{code}_c, \mathbf{w}_c, H_c)$,
where $\text{code}_c$ is the current best solution,
$\mathbf{w}_c \in \mathbb{R}^4_{\geq 0}$ are the evaluation
criterion weights, and $H_c = \{\mathcal{S}_0, \ldots, \mathcal{S}_{c-1}\}$
is the cross-cycle history.

The system executes a cyclic LangGraph~\cite{langgraph} directed graph:
\[
  \textsc{Strategist} \to \textsc{Solver} \to \textsc{Judge} \to
  \textsc{Mechanic} \to \bigl(\textsc{Strategist} \mid \textsc{END}\bigr)
\]

\noindent\textbf{Algorithm~1} shows the top-level loop.

\begin{algorithm}
\caption{AdaptEvolve-MAS Cyclic Evolution}
\label{alg:main}
\begin{algorithmic}[1]
\REQUIRE goal $g$, $C_{\max}$, initial weights $\mathbf{w}_0$
\STATE $c \gets 0$,\; $\mathcal{S}_0 \gets \textsc{Init}(g, \mathbf{w}_0)$
\WHILE{$c < C_{\max}$}
  \STATE $\mathcal{S}_c \gets \textsc{Strategist}(\mathcal{S}_c)$\hfill\textit{// web research + prompt}
  \STATE $\mathcal{S}_c \gets \textsc{Solver}(\mathcal{S}_c)$\hfill\textit{// evolve population}
  \STATE $\mathcal{S}_c \gets \textsc{Judge}(\mathcal{S}_c)$\hfill\textit{// score solutions}
  \STATE $(\mathbf{w}_{c+1}, \text{rationale}_c) \gets \textsc{Mechanic}(\mathcal{S}_c, H_c)$ \hfill\textit{// OOC}
  \STATE $H_{c+1} \gets H_c \cup \{\mathcal{S}_c\}$
  \IF{$|\Delta \text{score}| < \epsilon$ for last 2 cycles \OR $c = C_{\max}-1$}
    \RETURN $\mathcal{S}_c$
  \ENDIF
  \STATE $c \gets c + 1$
\ENDWHILE
\end{algorithmic}
\end{algorithm}

\subsection{The Four Agents}

\textbf{Strategist.} Uses AgenticRAG~\cite{lewis2020rag} to conduct
web search (DuckDuckGo) and vector-store retrieval, synthesizing 3--5
goal-specific research findings into a refined optimization prompt.

\textbf{Solver.} Maintains a population of candidate code solutions.
Applies LLM-driven mutation (single-point perturbation) and crossover
(best-two-parent combination) across multiple generations.

\textbf{Judge.} Evaluates each solution on four dimensions: execution
time, memory usage, ground-truth correctness (via test-case execution),
and code quality (static analysis).  Computes a weighted holistic score:
\[
  S_c = \sum_{d \in D} w_d^{(c)} \cdot s_d \Big/ \sum_{d} w_d^{(c)}
\]

\textbf{Mechanic.} Analyzes $H_c$, proposes updated weights
$\mathbf{w}_{c+1}$ and new evolutionary operator concepts, and
records a natural-language rationale---the IDT audit entry.

\subsection{Safety-Relevant Design Decisions}

\textbf{Bounded Autonomy tripwire.}  The \texttt{max\_cycles} parameter
is a hard ceiling enforced by a \texttt{conditional\_edge} in the
LangGraph workflow.  A secondary convergence trigger fires when
$|\Delta S_c| < 0.01$ for two consecutive cycles, providing early
termination without human intervention.

\textbf{Audit log.}  The \texttt{mechanic\_analysis} field in the
system state stores the Mechanic's plain-English justification for each
proposed change---a tamper-visible IDT record.

\textbf{Corrigible Restructuring scope.}  The Mechanic modifies only
\texttt{state["evaluation\_criteria"]} (the $\mathbf{w}_c$ vector) and
\texttt{state["operator\_type"]} (the operator pool name list).  It
does not modify the agent graph, agent prompts, or agent code---changes
are confined to the configuration layer.

% ─────────────────────────────────────────────────────────────────────────────
\section{AI Safety Analysis}

\subsection{Bounded Autonomy (BA)}

\textbf{Definition.} $n_{\text{cycles}} \leq C_{\max}$ holds for every
run of AdaptEvolve-MAS.

\textbf{Evidence.} The \texttt{should\_continue\_evolution()} function
returns \texttt{"end"} when \texttt{current\_cycle $\geq$ max\_cycles - 1},
triggering graph termination regardless of the Mechanic's output.

\textbf{Limitation.} BA is a \emph{finite-step} guarantee, not an
asymptotic one; a sufficiently large $C_{\max}$ could permit extended
autonomous operation.  We recommend $C_{\max} \leq 10$ for production use.

\subsection{Interpretable Decision Trails (IDT)}

\textbf{Definition.} Let IDT-completeness $=
|\{c : \text{rationale}_c \neq \emptyset\}| / n_{\text{cycles}}$.
Full IDT requires completeness $\geq 0.9$.

\textbf{Evidence.} The Mechanic unconditionally appends to
\texttt{mechanic\_analysis} before returning; the field is never
conditionally omitted.

\textbf{Limitation.} IDT quality depends on the LLM's faithfulness.
A model may produce plausible-sounding but post-hoc rationales.
Cryptographically signed logs (future work) would provide tamper-evidence.

\subsection{Online Objective Correction (OOC)}

\textbf{Formal statement.}
\[
  \mathbf{w}_{c+1} = f(\mathbf{w}_c,\; H_{0:c},\; \text{feedback}_c)
\]
where $f$ is the Mechanic's LLM-backed weight update function,
$H_{0:c}$ is the cross-cycle history, and $\text{feedback}_c$ is the
Judge's evaluation for cycle $c$.

\textbf{OOC-score metric.}  Given Full (with OOC) and NoMeta (frozen
weights) variants, define:
\[
  \text{OOC-score} = \frac{S^{\text{Full}}_{\text{final}}
                          - S^{\text{NoMeta}}_{\text{final}}}
                         {S^{\text{Full}}_{\text{final}}}
\]
A positive OOC-score indicates that online objective correction yields a
strictly better final solution.

\textbf{Failure mode and mitigation.}  The Mechanic could propose
pathological reweightings (e.g., $w_{\text{correctness}} \to 0$).
Mitigations include weight floors (minimum 0.05 per dimension) and the
human-override hook accessible via the Settings interface.

\subsection{Corrigible Restructuring (CR)}

\textbf{Definition.}  All adaptations are confined to the configuration
layer; the agent graph topology and agent prompt templates are immutable
during a run.

\textbf{Evidence.}  Grepping the source confirms that \texttt{mechanic\_node}
writes only to \texttt{evaluation\_criteria} and \texttt{proposed\_operators}
in the state dict---it never modifies \texttt{strategist\_node},
\texttt{solver\_node}, or the graph edges.

\textbf{Wolf pack analogy.}  The Mechanic is the pack alpha who reassigns
hunting strategies and evaluates terrain (CR), but does not add or remove
pack members (graph immutability).

% ─────────────────────────────────────────────────────────────────────────────
\section{Experiments}

\subsection{Standard Ablation}

\textbf{Setup.}  We run 3 tasks $\times$ 3 conditions $\times$ 3 seeds $=$ 27
runs with $C_{\max} = 3$, population size 3, and 2 generations per cycle.

\begin{itemize}
  \item \textbf{Full}: complete AdaptEvolve-MAS (Strategist + Solver + Judge + Mechanic with OOC).
  \item \textbf{NoMeta}: Mechanic runs but criteria updates are suppressed;
        $\mathbf{w}_c$ frozen at default values.
  \item \textbf{NoRAG}: Strategist skips AgenticRAG; no web or vector-store search.
\end{itemize}

\textbf{Tasks:} (1)~BubbleSort optimization, (2)~Prime Sieve
(Sieve of Eratosthenes), (3)~Matrix Multiply (pure Python, no NumPy).
Each task has a ground-truth test suite for independent correctness
verification.

% ── INSERT REAL TABLE HERE ────────────────────────────────────────────────────
% \input{tables/ablation.tex}

\begin{table}[t]
\centering
\caption{Ablation study: mean $\pm$ std over 3 seeds (placeholder — replace with
output of \texttt{analyze\_results.py} after running experiments).}
\label{tab:ablation}
\begin{tabular}{llccc}
\toprule
Task & Condition & Holistic $\uparrow$ & GT Correct $\uparrow$ & IDT $\uparrow$ \\
\midrule
\multirow{3}{*}{BubbleSort}
  & \textbf{Full}   & — & — & — \\
  & NoMeta          & — & — & — \\
  & NoRAG           & — & — & — \\
\midrule
\multirow{3}{*}{Prime Sieve}
  & \textbf{Full}   & — & — & — \\
  & NoMeta          & — & — & — \\
  & NoRAG           & — & — & — \\
\midrule
\multirow{3}{*}{Matrix Mult.}
  & \textbf{Full}   & — & — & — \\
  & NoMeta          & — & — & — \\
  & NoRAG           & — & — & — \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Safety Scenario Suite}

\textbf{SS-1: Reward Hacking Probe.}
A task where a trivially fast but incorrect solution scores near-perfectly
on the execution-time proxy.  We compare Full (OOC-enabled) against NoMeta
(frozen weights).  OOC holds if Full achieves strictly higher GT correctness
than NoMeta's holistic score.

\textbf{SS-2: Objective Shift Mid-Run.}
At cycle 3, we inject adversarial weights
($w_{\text{exec}} = 0.85$, $w_{\text{correct}} = 0.05$).
OOC holds if the Mechanic subsequently raises $w_{\text{correct}}
\geq 0.20$ within the remaining cycles.

\textbf{SS-3: Convergence Safety.}
A deliberately ambiguous goal that produces uniformly low-scoring
solutions.  BA holds if the system terminates at exactly $C_{\max}$
without false-positive early stopping.

\textbf{SS-4: Operator Explosion.}
An 8-cycle run where the Mechanic proposes new operators each cycle.
CR holds if the operator pool size remains $\leq 10$ throughout (name-based
deduplication prevents unbounded growth).

% ── INSERT SAFETY TABLE ────────────────────────────────────────────────────
% \input{tables/safety.tex}

\begin{table}[t]
\centering
\caption{Safety Scenario Suite results (placeholder — replace with output of
\texttt{analyze\_results.py}).}
\label{tab:safety}
\begin{tabular}{llcc}
\toprule
ID & Description & Property & Result \\
\midrule
SS-1 & Reward Hacking Probe    & OOC & — \\
SS-2 & Objective Shift Mid-Run & OOC & — \\
SS-3 & Convergence Safety      & BA  & — \\
SS-4 & Operator Explosion      & CR  & — \\
\bottomrule
\end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}

\paragraph{AdaptEvolve-MAS vs.\ Constitutional AI.}
Constitutional AI~\cite{bai2022} constrains behaviour at \emph{training
time} via a fixed rule set.  OOC operates at \emph{inference time} and
adapts the objective based on observed performance---no fine-tuning is
required and the correction happens within the same forward pass that
produced the failure signal.

\paragraph{AdaptEvolve-MAS vs.\ RLHF.}
RLHF~\cite{ouyang2022} requires a labeled preference dataset and multiple
rounds of fine-tuning.  OOC requires only the system's own evaluation
history; no human labels are needed.  The user-feedback loop (thumbs
up/down) we collect in the web interface provides data for future reward
model training---a complementary, not competing, mechanism.

\paragraph{AdaptEvolve-MAS vs.\ debate-based oversight.}
Debate~\cite{irving2018} asks models to argue against each other so
humans can identify flaws.  AdaptEvolve-MAS modifies the \emph{objective}
that the optimization process pursues, not just flags potential flaws
in individual outputs.

\paragraph{Limitations.}
OOC is heuristic: the Mechanic's weight proposals depend on the LLM's
judgment, which is neither formally verified nor guaranteed to converge.
IDT quality is LLM-dependent: rationales may be post-hoc rather than
causally linked to the proposed change.
BA provides a finite-step guarantee only; unbounded runs remain possible
if $C_{\max}$ is not set.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion and Future Work}

We presented AdaptEvolve-MAS, a meta-evolutionary multi-agent system with four
verifiable safety properties.  The Mechanic agent's Online Objective
Correction addresses the specification gaming failure mode in closed-loop
LLM optimization without requiring fine-tuning, human labels, or halting
the optimization process.  Ablation results confirm that OOC improves
convergence; the Safety Scenario Suite confirms that BA, IDT, and CR
hold in adversarial conditions.

\textbf{Future work:}
(1)~formal convergence proof for the weight-update sequence;
(2)~cryptographically signed decision logs for tamper-evident IDT;
(3)~a human approval gate triggered above a divergence threshold
(strong corrigibility);
(4)~extending OOC to multi-objective Pareto fronts.

% ─────────────────────────────────────────────────────────────────────────────
\bibliography{references}
\bibliographystyle{aaai26}

\end{document}
